Measurement Bias from Address Aliasing

### 2014-05-02 paper.tex
Update introduction and methodology
Fixes based on feedback

	% Bad motivation start:
    % In this paper we show how an optimization in the out-of-order execution engine on modern Intel microarchtectures can affect performance depending on run-time memory layout. 
    In this paper we identify the underlying mechanics resulting in measurement bias on modern Intel microarchtectures. 
    The magnitude of this performance bias has been shown to be significant, and can be as high as a factor two.

@@

Measurement bias has been shown to be commonplace and significant in real applications [move cite here?].
Previous work looks at environmental factors such as the size of Unix environment variables, which affects memory by offsetting the initial position of stack.
Because of the [enormous] complexity of modern hardware, it is often impossible to predict exactly how seemingly innocuous things such as environment variables ultimately affect program execution.
[In fact], the best one can do to accurately assess performance is generally to average over a large sample set of execution contexts [not selling our case here]. \cite{Mytkowicz:2008:OE&MB, Mytkowicz:2009:WrongData, Mytkowicz:2008:Easy}

Effects of bias has been shown to be commonplace and significant, yet unpredictable and difficult to deal with [cite/ref].
Previous work focuses mostly of how to account for bias by proper sampling, using randomization of execution contexts and statistical methods.
Our approach is instead to try to understand some of the underlying reasons to why bias can occur.
Usually, the root explanation will be that different memory layouts interact with various hardware buffers.
Certain mechanisms such as cache contention or branch misprediction are already well understood.
[We collect 200+ events using HW counters, and search for correlations.]
[In this paper,] we show how a much lesser known, presumably Intel specific, optimization on the out-of-order execution pipeline can [be the] cause of measurement bias.
Analyzing a program that exhibits strong bias from environment size, we find a correlation with resource stalls from \emph{address aliasing}.
In short, the CPU uses a [conservative] heuristic for determining whether memory accesses can be retired out of order, comparing only the last 12 virtual address bits.
False dependencies incur performance penalties, which can explain many common cases of measurement bias.

In addition to environment variables, we look at dynamic heap allocation libraries as an external bias trigger [amplifier?].
We find that [Heap allocation turns out to be particularly interesting for the following reason:] most libraries tend to produce worst case behavior by default with respect to aliasing.
We show that the cost of address aliasing can be significant, exemplified with a simple program yielding more than 50~\% speedup based on heap address alignment alone.
[This next part might be added to the abstract in some form:] 
Fortunately, with an understanding of the underlying mechanisms, measurement bias from address aliasing can often be quite accurately predicted.
For some applications it might even make sense for programmers to actively take aliasing into account for architecture specific optimization, as neither compilers nor heap allocators seem to handle it well.

@@

Performance counters are supported in the Linux kernel, and we access them via a tool called \emph{perf}. 
We use the \texttt{perf-stat} command in all experiments, which accepts raw event codes listed in the Intel reference manual \cite{Intel:2013:Volume3B}.
A small python script is used to collect an exhaustive set of all available counters, which amounts to about 200 on our architecture. [Might add "and environmental sizes".]
Interesting events are identified [isolated] by computing correlation to cycle count, measuring all counters over a series of execution contexts.
[Both lines overlap:]
Cache related metrics are monitored in order to rule out cache as the underlying cause of bias, such as hit rates of [micro-ops] for each level of cache~\cite{Intel:2012:OptimizationManual}.
[Metrics related to cache, such as hit rates for different levels, are closely monitored in order to exclude cache as the underlying cause of bias.]

Results are averaged [not median?] over multiple runs, and all benchmarks are done on a machine under minimal load.
Our setup consists of an Intel® Core™ i7-4770K CPU @ 3.50 GHz, 32 GiB DDR3 1600 MHz memory, running a 64-bit installation of Ubuntu 13.10.
Because bias can be hardware dependent, and to keep the scope of this work manageable, we choose to focus only on the Intel ``Haswell'' microarchitecture specifically.
Code samples are compiled using the GCC toolchain, version 4.8.1 [Perhaps the full name? 4.4.6-11ubuntu2].

[Add a link to a paper that lists good practice?: ] 
We also have to ensure we are not affected by measurement bias beyond what we are actually trying to observe.
Most importantly, this means keeping the memory address space under control.
For security reasons, addresses of stack, heap and dynamic libraries are often randomized at load time, a technique known as Address Space Layout Randomization.\cite{Shackham:2004:ASLR} % Should maybe have more generic reference
By disabling \emph{ASLR}, we are able to execute the same program multiple times with identical virtual address spaces.
Bias from environment size is provoked by setting a dummy environment variable to $n$ number of zero characters, starting from a minimal environment.\footnote{Because perf-stat itself adds a few variables, the environment will never be completely empty.}

\emph{Frequency scaling} is disabled, keeping the processor's clock speed fixed.
Finally, we disable \emph{Hyper-threading}, which reduces the possibility for resource contention [in the CPU] between threads.

>>>>

    In this paper we identify the underlying mechanics resulting in measurement bias on modern Intel microarchtectures. 
    The magnitude of this performance bias has been shown to be significant, and can be as high as a factor two.

@@

% This is basically former background section
Measurement bias has been shown to be commonplace and significant in real applications \cite{Mytkowicz:2008:OE&MB, Mytkowicz:2009:WrongData, Mytkowicz:2008:Easy}.
Previous work looks at environmental factors such as the size of Unix environment variables, which affects memory by offsetting the initial position of stack.
Because of the complexity of modern hardware, it is often impossible to predict exactly how seemingly innocuous things such as environment variables ultimately affect program execution.
While effects of bias has been shown to be commonplace and significant, they are unpredictable and difficult to deal with.
Without an accurate understanding of the underlying memory interactions, the best one can do to accurately assess performance is generally to average over a large sample set of execution contexts.
Previous work focuses mostly of how to account for bias by proper sampling, using randomization of execution contexts and statistical methods. % Cite is basically the same as above, all three are about the same -- difficult to split.

Our approach is instead to try to understand some of the underlying reasons to why bias can occur, in an attempt to utilize this knowlege for potential optimization. % Could have a sentence with cite of Blind Optimization.
Usually, the root explanation will be that different memory layouts interact with various hardware buffers.
Certain mechanisms such as cache contention or branch misprediction are already well understood, thus we are looking for the more esoteric processor features.
Analyzing a program that exhibits strong bias from environment size, collecting about 200 events using hardware performance counters, we find strong correlation with resource stalls from \emph{address aliasing}.
Aliasing is an artifact of a much lesser known, presumably Intel specific, optimization on the out-of-order execution pipeline.
In short, the CPU uses a heuristic for determining whether memory loads are dependent on previous writes, comparing only the last 12 virtual address bits.
False dependencies incur performance penalties, and we show how this can explain many common cases of measurement bias.

In addition to environment variables, we look at dynamic heap allocation libraries as an external factor that can trigger (or amplify) bias resulting from address aliasing.
We find that most libraries tend to produce worst case behavior by default with respect to aliasing, caused by favoring page alignment for large allocations.
The cost of address aliasing can be significant, exemplified with a simple program yielding more than 50~\% speedup based on heap address alignment alone.
% [This next part might be added to the abstract in some form:] 
Fortunately, with an understanding of the underlying mechanisms, measurement bias from address aliasing can often be quite accurately predicted.
For some applications it might even make sense for programmers to actively take aliasing into account for architecture specific optimization, as neither compilers nor heap allocators seem to handle it well.

@@

Performance counters are supported in the Linux kernel, and we access them via a tool called \emph{perf}. 
We use the \texttt{perf-stat} command in all experiments, which accepts raw event codes listed in the Intel reference manual \cite{Intel:2013:Volume3B}.
A small python script is used to collect an exhaustive set of all available counters, which amounts to about 200 on our architecture.
The same script simulates changes to environment size, which is provoked by setting a dummy environment variable to $n$ number of zero characters, starting from a minimal environment.\footnote{Because perf-stat itself adds a few variables, the environment will never be completely empty.}
Interesting events are identified by computing linear correlation to cycle count, measuring all counters over a series of execution contexts.
Cache related metrics are monitored in order to rule out cache as the underlying cause of bias, such as hit rates of micro-ops for each level of cache~\cite{Intel:2012:OptimizationManual}.
Results are averaged over multiple runs to reduce potential random error, supplying the \texttt{-r} option to perf-stat. % Not median, good question. man perf-stat says average + std.dev.

Our setup consists of an Intel\textregistered{} Core\texttrademark{} i7-4770K CPU @ 3.50 GHz, running 64-bit Ubuntu 13.10.
Because bias can be hardware dependent, and to keep the scope of this work manageable, we choose to focus only on this processor and the Intel ``Haswell'' microarchitecture specifically.
Code samples are compiled using the GCC toolchain, version 4.8.1-10ubuntu9.

We also have to ensure we are not affected by measurement bias beyond what we are actually trying to observe, and in general follow best practices on gathering data \cite{Mytkowicz:2009:WrongData}.
Most importantly, this means keeping the memory address space under control.
For security reasons, addresses of stack, heap and dynamic libraries are often randomized at load time, a technique known as Address Space Layout Randomization.\cite{Shackham:2004:ASLR} % Should maybe have more generic reference
By disabling \emph{ASLR}, we are able to execute the same program multiple times with identical virtual address spaces.
All benchmarks are done on a machine under minimal load, and \emph{Frequency scaling} is disabled to keep the processor's clock speed fixed.
Finally, we disable \emph{Hyper-threading}, which reduces the possibility for resource contention between threads.
